General limits

### 2017-09-18 papers/talk-7/paper-7.tex
Fixed all explanations and notations. Should be good to submit now

%% TODO: consistent notation, DI D(I) D_I, also for maps D_m Dm

@@

The functor $D$ yields
a collection of objects and morphisms in the category $\mathscr C$
that looks just like $\bm I$.
We also say $D$ \textbf{identifies} diagrams of shape $\bm I$.
The shape category has no further meaning. Its only purpose
is to define how the diagram is supposed to look like.
The reason we use an intermediate
category $\bm I$ to define the shape is that
there might be many parts in $\mathscr C$ which have the shape $\bm I$,
especially when $\bm I$ is so small.
We therefore get all different diagrams of that shape by considering
all different functors $D$.
Diagrams do not necessarily have to be small. Their shape can even be infinite.
However most interesting diagrams tend to be very small.\\
We are typically interested in diagrams of a specific category,
$\mathscr C$ in the example. Therefore we usually leave out $\bm I$ and $D$,
and we just draw the identified part of $\mathscr C$ instead.\\
The next part on the way to the limit definition is the concept of a cone which
is what we address now.
\begin{definition}
  Let $D \from \bm{I} \to \mathscr{C}$ be a diagram.
  A \textbf{cone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $U \overset{f_I}{\rightarrow} D_I, \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      & U \arrow[dl, swap, "f_I"] \arrow{dr}{f_J}&\\
      D_I \arrow[rr, swap, "D_m"] & & D_J\\
    \end{tikzcd}
  \]
  That is $f_J = D(m) \circ f_I$.\\
  For a cone we write $(U \overset{f_I}{\to} D(I))_{I\in \bm{I}}$.
  We call $U$ the \textbf{vertex}, and $f_I$ the \textbf{projection maps} of the cone.
\end{definition}
\begin{remark}
  When it is clear from the context that $U$ is the vertex of a cone, then we use just
  the vertex $U$
  to refer to the whole cone.\\
  Note that the underlying diagram determines how the cone looks like.
\end{remark}
%% TODO: Notation DI, D(I), D_I ??
\begin{example}
  A cone over the diagram of pullbacks
  (\hyperref[diagramexample]{Example \ref*{diagramexample}}) looks like this.
  \[
    \begin{tikzcd}
      U \arrow[bend right=30]{ddr}{f_I} \arrow[bend left=30]{drr}{f_k} \arrow[bend left=20, swap]{ddrr}{f_J}&   &\\
      & & D_K \arrow{d}{D_n} \\
      & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The commuting condition says that $D_m \circ f_I = f_J$ and $D_n \circ f_K = f_J$.
\end{example}

@@

\begin{definition}
  A cone $(U \overset{f_I}{\to} D(I))_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cones $(V \overset{g_I}{\to} D(I))_{I\in \bm{I}}$ there is a unique
  map $h \from V \to U$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      V \arrow[dr, swap, "g_I"] \arrow{r}{h} & U \arrow{d}{f_I} \\
      & D_I
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}

Intuitively, a cone is universal if it
is universally the ``closest'' to the diagram.

\begin{example}
  This is an example of a universal cone over the
  diagram of pullbacks
  (\hyperref[diagramexample]{Example \ref*{diagramexample}}).
    \[
    \begin{tikzcd}
      {\color{red}V}
      \arrow[bend right=40,color=red]{dddrr}{g_I}
      \arrow[bend left=40, color=red]{ddrrr}{g_K}
      \arrow[bend left=30, color=red, swap]{dddrrr}{g_J}
      \arrow[dashed,swap]{dr}{h}
      & & & \\
      & {\color{blue}U}
      \arrow[bend right=30,color=blue]{ddr}{f_I}
      \arrow[bend left=30, color=blue]{drr}{f_K}
      \arrow[bend left=20, color=blue, swap]{ddrr}{f_J}&   &\\
      & & & D_K \arrow{d}{D_n} \\
      & & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The universal cone {\color{blue}U}
  is drawn in blue and the other cone {\color{red}V} is drawn in red.
  Both cones obviously have to satisfy the commuting condition for cones.
  Additionally the commuting condition for universal cones states that
  $g_I = f_I \circ h$, $g_J = f_J \circ h$, and $g_K = f_K \circ h$.
  Therefore this example commutes in its entirety.\\
  Note that there could be several other cones like $V$.
\end{example}

@@

\begin{lemma}
  \label{finiteproducts}
  Let $\mathscr C$ be a category with a terminal object $1$ and binary products.
  Then we can construct all finite products.
\end{lemma}
\begin{proof}
  \begin{itemize}
  \item The empty product is just the terminal object by
    \hyperref[emptyproduct]{Example \ref*{emptyproduct}}
  \item The unary product of any object $A$ is constructed by the binary product
    $A \times A$.
    We have $1_A \from A \to A$, and $\pi_1, \pi_2 \from A \times A \to A$.
    By the universal property of products there exists a unique map $h \from A \to A \times A$.
    We have $1_A = \pi_1 \circ h$.\\
    On the other hand we have $(\pi_1 \circ h) \circ \pi_1 = 1_a \circ \pi_1 = \pi_1 \circ ( h \circ \pi_1)$,
    and therefore $1_A = h \circ \pi_1$.
    We have constructed an isomorphism and therefore $A \times A \iso A$.
  \item The binary product exists by assumption for all objects $A, B$.
  \item We want to build some finite product over objects $D_1, \dots D_k$.
    Assume by induction that the following two products exist.
    \[ P_1 \coloneqq \displaystyle \prod_{i \in \{1, \dots, m \} } D_i \text{ , and } P_2 \coloneqq \displaystyle \prod_{i \in \{m+1, \dots, k\} } D_i \]
    with projection maps $(\pi_i)_{i \in \{1, \dots, m \} }$ for $P_1$, and $(\tau_i)_{i \in \{m+1, \dots, k\} }$ for $P_2$.
    Then there exists a binary product of $P_1 \times P_2$ with projection maps $f_{P_1}, f_{P_2}$.
    $P_1 \times P_2$ has maps $\pi_i \circ f_{P_1}$ and $\tau_i \circ f_{P_2}$.
    $P_1 \times P_2$ is therefore a cone.
    Since $P_1$ and $P_2$ are products, any other cone $V$ over them has unique maps $h_{P_1}$ and $h_{P_2}$.
    The unique morphism $V \rightarrow P_1 \times P_2$ is given by $\langle h_{P_1}, h_{P_2} \rangle$.
    $P_1 \times P_2$ is therefore universal.
  \end{itemize}
  Intuitively the construction works by stacking binary products on top of each other.
  Since we assume $\mathscr C$ to have all products, the product of two products exists. This way we can create
  arbitrarily large finite products.
\end{proof}

@@

\begin{theorem}
  \label{pullbackconstruction}
  Let $\mathscr C$ be a category with finite products and equalizers.
  Then we can construct pullbacks.
\end{theorem}
\begin{proof}
We start with the diagram for pullbacks.
\[
  \begin{tikzcd}
    & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
We need to construct a limit of this diagram.
First we take the binary product of $I$ and $K$ which exists by assumption.
\[
  \begin{tikzcd}
    I \times K \arrow{r}{p_K} \arrow{d}{p_I} & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
This looks already similar to a pullback.
However, we need to ensure the commuting condition.
For this purpose we use an equalizer which also exists by assumption.
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    I \times K
    \arrow[r,shift right,swap, "m \circ p_I"] \arrow[r,shift left, "n \circ p_K"]
    & J \\
  \end{tikzcd}
\]
We now unfold this drawing.
\[
  \begin{tikzcd}
    E \arrow{dr}{e}& &\\
    & I \times K \arrow[swap]{d}{p_I} \arrow{r}{p_K}& K \arrow{d}{n}\\
    & I \arrow[swap]{r}{m} & J
  \end{tikzcd}
  \qquad \bm{\rightsquigarrow} \qquad 
  \begin{tikzcd}
    E \arrow[swap]{d}{p_I \circ e} \arrow{r}{p_K \circ e} & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
$E$ satisfies the commuting and universality conditions of limits
since it is already an equalizer.
%% TODO: more detail...
Therefore it is a pullback.
\end{proof}

\begin{fact}
  \label{pullbackproduct}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct finite products.
\end{fact}
\begin{proof}
Finite products can be constructed
with the terminal object and binary products by
\hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Therefore we only show how to construct binary products.
Consider following the diagram:
\[
  \begin{tikzcd}
    & B \arrow{d} \\
    A \arrow{r} & 1
  \end{tikzcd}
\]
The pullback $U$ exists by assumption.
\[
  \begin{tikzcd}
    U \arrow{r}{p_B} \arrow{d}{p_A} & B \arrow{d}{f} \\
    A \arrow{r}{g} & 1
  \end{tikzcd}
\]
Since $U$ is a pullback we have for any other cone $V$ with projections $q_A, q_B$
a unique map $h \from V \to U$, such that
$q_A = p_A \circ h$, and $q_B = p_B \circ h$, which is exactly the commuting condition
for products.
\end{proof}

Now that we know that a category with pullbacks and a terminal object
also has finite products, we can proceed to construct equalizers as well
as their construction build on the existence of products in this category.

\begin{theorem}
  \label{pullbackequalizer}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct equalizers.
\end{theorem}
\begin{proof}
  We start with the initial diagram for equalizers.
  \[
    \begin{tikzcd}
      A \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} & B
    \end{tikzcd}
  \]
  We transform this into the following diagram.
  \[
    \begin{tikzcd}
      & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  The product $B \times B$ exists by \hyperref[pullbackproduct]{Theorem \ref*{pullbackproduct}}.
  Furthermore we have $B \times B \iso B$ by \hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
  The above diagram is therefore equivalent.\\
  We then have the pullback $E$, such that the following diagram commutes.
  \[
    \begin{tikzcd}
      E \arrow{r}{h} \arrow{d}{e} & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  Since $E$ is a pullback the square above commutes and
  we get $f \circ e = 1_B \circ h = h = 1_B \circ h = g \circ e$, and $E$ is therefore
  indeed an equalizer of $f$ and $g$.
\end{proof}

@@

\begin{theorem}
  \label{arbitrarylimits}
  Let $\mathscr C$ be a category that has all (finite) products and equalizers.
  Let $D : \bm I \to \mathscr C$ be a (finite) diagram.
  Then we can construct a limit of $D$.
\end{theorem}
\begin{proof}
We use the following construction:
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    \displaystyle \prod_{I \in \bm{I}} D(I)
    \arrow[r,shift left, "s"] \arrow[r,shift right,swap,"t"] &
    \displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D(J)
  \end{tikzcd}
\]
with the maps $s, t$ that have components $s_n, t_n$ for all $K \overset{n}{\to} J \in \bm{I}$:
\begin{align*}
  s_n &= Dn \circ p_K\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_K}{\rightarrow} D_K \overset{Dn}{\rightarrow} D_J \\
  t_n &= p_J\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_J}{\rightarrow} D_J
\end{align*}

The product $\displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D(J)$ has components for each map in the diagram.
We take the diagram for pullbacks as an example for clarification:
\begin{tikzcd}
 I \arrow{r}{m} & J & \arrow[swap]{l}{n} K
\end{tikzcd}.\\
For this example diagram the product has the following projections. One for each identity:
$p_{1_I} \from \prod \to I,\ p_{1_J} \from \prod \to J,\ p_{1_K} \from \prod \to K$,
and one for each remaining map in the diagram: $p_m \from \prod \to J,\ p_n \from \prod \to J$.\\
All products in the above construction exist by assumption.
We take $E$ as the equalizer of $s$ and $t$ which also exists by assumption.
We therefore have $s \circ e = t \circ e$.
The maps $s$ and $t$ have components. The equality has to hold for all
components and we get
\begin{align*}
  s \circ e &= t \circ e\\
  s_n \circ e &= t_n \circ e, \quad \forall\ K \overset{n}{\rightarrow} J \in \bm I \\
  Dn \circ p_K \circ e &= p_J \circ e
\end{align*}
The goal is to construct a limit.
We do this by constructing a cone and showing that it is universal.
For the construction of a cone
we need a vertex and projection maps into every object of the diagram,
such that the commuting condition of cones is satisfied.
We take $E$ as the vertex and
$\uprho_I \coloneqq p_I \circ e, \forall\ I \in \bm I$
as the projection maps of the cone.
Clearly, we have all necessary projection maps and we only need to verify
the commuting condition of cones.
We have all maps $K \overset{n}{\rightarrow} J \in \bm I$
\begin{align*}
  Dn \circ p_K \circ e &= p_J \circ e \\
  Dn \circ \uprho_K &= \uprho_J
\end{align*}
which corresponds to the commuting diagram
\[
  \begin{tikzcd}
    E \arrow{r}{\uprho_K} \arrow{dr}[swap]{\uprho_J} & D_K \arrow{d}{Dn}\\
    & D_J
  \end{tikzcd}
\]
This is exactly the commuting condition we need for a cone.\\
Since any other cone over $D$
is also a fork of $s$ and $t$
it follows that the equalizer $E$ is also a universal cone
and therefore a limit of $D$.
\end{proof}
\begin{remark}
  If $\mathscr C$ has all products including infinite products then
  this construction works even for infinite diagrams.
  If $\mathscr C$ only has finite products, then we can
  use this construction for
  limits of finite diagrams.
\end{remark}

To summarize, we can construct arbitrary limits in the following way.
If $\mathscr C$ has all products and equalizers then we can
construct all limits by
\hyperref[arbitrarylimits]{Theorem \ref*{arbitrarylimits}}.
Equalizer can be constructed using finite products and pullbacks
by \hyperref[arbitrarylimits]{Theorem \ref*{pullbackequalizer}}.
Finite products can in turn be constructed from binary products and a terminal object
by \hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Binary products can be constructed using a pullback and a terminal object
by \hyperref[arbitrarylimits]{Theorem \ref*{pullbackproduct}}.
Having pullbacks and a terminal object is therefore the lowest requirement from
which it is possible to bootstrap the constructions of other limit instances
and even of arbitrary limits.

>>>>

@@

The functor $D$ yields
a collection of objects and morphisms in the category $\mathscr C$
that looks just like $\bm I$.
We also say $D$ \textbf{identifies} diagrams of shape $\bm I$.
The shape category has no further meaning. Its only purpose
is to define how the diagram is supposed to look like.
The reason we use an intermediate
category $\bm I$ to define the shape is that
there might be many parts in $\mathscr C$ which have the shape $\bm I$,
especially when $\bm I$ is so small.
We therefore get all different diagrams of that shape by considering
all different functors $D$.
Diagrams do not necessarily have to be small. Their shape can even be infinite.
However most interesting diagrams tend to be very small.\\
We are typically interested in diagrams of a specific category,
$\mathscr C$ in the example. Therefore we usually leave out $\bm I$ and $D$,
and we just draw the identified part of $\mathscr C$ instead.\\
The next part on the way to the limit definition is the concept of a cone which
is what we address now.
\begin{definition}
  Let $D \from \bm{I} \to \mathscr{C}$ be a diagram.
  A \textbf{cone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $U \overset{f_I}{\rightarrow} D_I, \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      & U \arrow[dl, swap, "f_I"] \arrow{dr}{f_J}&\\
      D_I \arrow[rr, swap, "D_m"] & & D_J
    \end{tikzcd}
  \]
  That is, $f_J = D_m \circ f_I$.\\
  For a cone we write $(U \overset{f_I}{\to} D_I)_{I\in \bm{I}}$.
  We call $U$ the \textbf{vertex}, and $f_I$ the \textbf{projection maps} of the cone.
\end{definition}
\begin{remark}
  When it is clear from the context that $U$ is the vertex of a cone, then we use just
  the vertex $U$ to refer to the whole cone.
\end{remark}
\begin{example}
  A cone over the diagram of pullbacks
  (\hyperref[diagramexample]{Example \ref*{diagramexample}}) looks like this.
  \[
    \begin{tikzcd}
      U \arrow[bend right=30]{ddr}{f_I} \arrow[bend left=30]{drr}{f_K} \arrow[bend left=20, swap]{ddrr}{f_J}&   &\\
      & & D_K \arrow{d}{D_n} \\
      & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The commuting condition says that $D_m \circ f_I = f_J$ and $D_n \circ f_K = f_J$.
\end{example}

@@

\begin{definition}
  A cone $(U \overset{f_I}{\to} D_I)_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cones $(V \overset{g_I}{\to} D_I)_{I\in \bm{I}}$ there is a unique
  map $h \from V \to U$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      V \arrow[dr, swap, "g_I"] \arrow{r}{h} & U \arrow{d}{f_I} \\
      & D_I
    \end{tikzcd}
  \]
  That is, $g_I = f_I \circ h$.
\end{definition}
Intuitively, a cone is universal if it
is universally the ``closest'' to the diagram.

\begin{example}
  This is an example of a universal cone over the
  diagram of pullbacks.
  \[
    \begin{tikzcd}
      {\color{red}V}
      \arrow[bend right=40,color=red]{dddrr}{g_I}
      \arrow[bend left=40, color=red]{ddrrr}{g_K}
      \arrow[bend left=30, color=red, swap]{dddrrr}{g_J}
      \arrow[dashed,swap]{dr}{h}
      & & & \\
      & {\color{blue}U}
      \arrow[bend right=30,color=blue]{ddr}{f_I}
      \arrow[bend left=30, color=blue]{drr}{f_K}
      \arrow[bend left=20, color=blue, swap]{ddrr}{f_J}&   &\\
      & & & D_K \arrow{d}{D_n} \\
      & & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The universal cone {\color{blue}U}
  is drawn in blue and the other cone {\color{red}V} is drawn in red.
  Both cones obviously have to satisfy the commuting condition for cones.
  Additionally the commuting condition for universal cones states that
  $g_I = f_I \circ h$, $g_J = f_J \circ h$, and $g_K = f_K \circ h$.
  Therefore this example commutes in its entirety.\\
  Note that there could be several other cones like $V$.
\end{example}

@@

\begin{lemma}
  \label{finiteproducts}
  Let $\mathscr C$ be a category with a terminal object $1$ and binary products.
  Then we can construct all finite products.
\end{lemma}
\begin{proof}
  The idea is to first show the existence of the empty and the unary product.
  For larger products we stack binary products on top of each other.
  \begin{itemize}
  \item The empty product is just the terminal object by
    \hyperref[emptyproduct]{Example \ref*{emptyproduct}}
  \item The unary product of any object $A$ is constructed from the binary product
    $A \times 1$.
    We have maps $1_A \from A \to A$, $\pi_A \from A \times 1 \to A$, and $\pi_1 \from A \times 1 \to 1$.
    Every other object $B$ has a unique map $t_B \from B \to 1$. If $B$ has a map $f \from B \to A$, then
    it factors uniquely through $A \times 1$ by the universal property of products.
  \item The binary product exists by assumption for all objects $A, B$.
  \item We build some finite product over objects $D_1, \dots D_k$.
    Assume by induction that the following two smaller products exist.
    \[ P_1 \coloneqq \displaystyle \prod_{i \in \{1, \dots, \frac{k}{2} \} } D_i \text{ , and } P_2 \coloneqq \displaystyle \prod_{i \in \{\frac{k}{2} + 1, \dots, k\} } D_i \]
    with projection maps $(\pi_i)_{i \in \{1, \dots, \frac{k}{2} \} }$ for $P_1$, and $(\tau_i)_{i \in \{\frac{k}{2}+1, \dots, k\} }$ for $P_2$.
    Then there exists a binary product of $P_1 \times P_2$ with projection maps $f_{P_1}, f_{P_2}$.
    $P_1 \times P_2$ has maps $\pi_i \circ f_{P_1}$ and $\tau_i \circ f_{P_2}$.
    $P_1 \times P_2$ is therefore a cone.
    Since $P_1$ and $P_2$ are products, any other cone $V$ over them has unique maps $h_{P_1}$ and $h_{P_2}$.
    The unique morphism $V \rightarrow P_1 \times P_2$ is given by $\langle h_{P_1}, h_{P_2} \rangle$.
    $P_1 \times P_2$ is therefore universal.
  \end{itemize}
\end{proof}

@@

\begin{theorem}
  \label{pullbackconstruction}
  Let $\mathscr C$ be a category with finite products and equalizers.
  Then we can construct pullbacks.
\end{theorem}
\begin{proof}
We start with the diagram for pullbacks.
\[
  \begin{tikzcd}
    & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
We need to construct a limit of this diagram.
First we take the binary product of $I$ and $K$ which exists by assumption.
\[
  \begin{tikzcd}
    I \times K \arrow{r}{p_K} \arrow{d}{p_I} & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
This looks already similar to a pullback.
However, we need to ensure the commuting condition.
For this purpose we use an equalizer which also exists by assumption.
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    I \times K
    \arrow[r,shift right,swap, "m \circ p_I"] \arrow[r,shift left, "n \circ p_K"]
    & J \\
  \end{tikzcd}
\]
We now unfold this drawing.
\[
  \begin{tikzcd}
    E \arrow{dr}{e}& &\\
    & I \times K \arrow[swap]{d}{p_I} \arrow{r}{p_K}& K \arrow{d}{n}\\
    & I \arrow[swap]{r}{m} & J
  \end{tikzcd}
  \qquad \bm{\rightsquigarrow} \qquad 
  \begin{tikzcd}
    E \arrow[swap]{d}{p_I \circ e} \arrow{r}{p_K \circ e} & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
Since $E$ is an equalizer we have $e \circ p_K \circ n = e \circ p_I \circ m$.
Therefore, $E$ satisfies the commuting condition for cones over a pullback square.
Furthermore note that any other cone $V$ over this pullback square has maps $f \from V \to I$,
$g \from V \to K$.
By the universal property of products there is a unique map $m \from V \to I \times K$.
Since $E$ is an equalizer, there is a unique map $h \from V \to E$,
and $E$ is indeed a universal cone, that is, a pullback.
\end{proof}

\begin{fact}
  \label{pullbackproduct}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct finite products.
\end{fact}
\begin{proof}
Finite products can be constructed
with the terminal object and binary products by
\hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Therefore we only show how to construct binary products.
Consider following the diagram:
\[
  \begin{tikzcd}
    & B \arrow{d} \\
    A \arrow{r} & 1
  \end{tikzcd}
\]
The pullback $U$ exists by assumption.
\[
  \begin{tikzcd}
    U \arrow{r}{p_B} \arrow{d}{p_A} & B \arrow{d}{f} \\
    A \arrow{r}{g} & 1
  \end{tikzcd}
\]
Every object has a unique map to $1$ because it is terminal.
The cones over $A$ and $B$ are therefore exactly the cones
over $A$, $B$, and $1$.
Since $U$ is a pullback, it is therefore also the product.
\end{proof}

Now that we know that a category with pullbacks and a terminal object
also has finite products, we can proceed to construct equalizers as well.
Their construction builds on the existence of products in this category.

\begin{theorem}
  \label{pullbackequalizer}
  Let $\mathscr C$ be a category with pullbacks and a terminal object $1$.
  Then we can construct equalizers.
\end{theorem}
\begin{proof}
  We start with the initial diagram for equalizers.
  \[
    \begin{tikzcd}
      A \arrow[shift left]{r}{f} \arrow[shift right, swap]{r}{g} & B
    \end{tikzcd}
  \]
  To construct the equalizer of $f$ and $g$ we take the pullback of the following diagram
  \[
    \begin{tikzcd}
      & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  The product $B \times B$ exists by \hyperref[pullbackproduct]{Theorem \ref*{pullbackproduct}}.
  We then have the pullback $E$, such that the following square commutes.
  \[
    \begin{tikzcd}
      E \arrow{r}{h} \arrow{d}{e} & B \arrow{d}{\langle 1_B, 1_B \rangle} \\
      A \arrow{r}{\langle f,g \rangle} & B \times B
    \end{tikzcd}
  \]
  The product $B \times B$ has projection maps $\pi_1, \pi_2$.
  By the commuting condition we get
  \begin{align*}
    & f \circ e \\
    =\ & \pi_1 \circ \langle f, g \rangle \circ e\\
    =\ & \pi_1 \circ \langle 1_B, 1_B \rangle \circ h\\
    =\ & \pi_2 \circ \langle 1_B, 1_B \rangle \circ h\\
    =\ & \pi_2 \circ \langle f,g \rangle \circ e\\
    =\ & g \circ e
  \end{align*}
  Therefore, $E$ satisfies the commuting condition for equalizers.
  Universality follows, since any other fork of $f$ and $g$ is also a pullback
  cone over the above pullback square.
\end{proof}

@@

\begin{theorem}
  \label{arbitrarylimits}
  Let $\mathscr C$ be a category that has all (finite) products and equalizers.
  Let $D : \bm I \to \mathscr C$ be a (finite) diagram.
  Then we can construct a limit of $D$.
\end{theorem}
\begin{proof}
We use the following construction:
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    \displaystyle \prod_{I \in \bm{I}} D_I
    \arrow[r,shift left, "s"] \arrow[r,shift right,swap,"t"] &
    \displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D_J
  \end{tikzcd}
\]
with the maps $s, t$ that have components $s_n, t_n$ for all $K \overset{n}{\to} J \in \bm{I}$:
\begin{align*}
  s_n &= Dn \circ p_K\ \hat{=} \prod_{I \in \bm{I}} D_I
        \overset{p_K}{\rightarrow} D_K \overset{Dn}{\rightarrow} D_J \\
  t_n &= p_J\ \hat{=} \prod_{I \in \bm{I}} D_I
        \overset{p_J}{\rightarrow} D_J
\end{align*}

The product $\displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D_J$ has components for each map in the diagram.
We take the diagram for pullbacks as an example:
\begin{tikzcd}
 I \arrow{r}{m} & J & \arrow[swap]{l}{n} K
\end{tikzcd}.
For this example diagram the product has the following projections. One for each identity:
$p_{1_I} \from \prod \to I,\ p_{1_J} \from \prod \to J,\ p_{1_K} \from \prod \to K$,
and one for each remaining map in the diagram: $p_m \from \prod \to J,\ p_n \from \prod \to J$.\\
All products in the above construction exist by assumption.
We take $E$ as the equalizer of $s$ and $t$ which also exists by assumption.
We therefore have $s \circ e = t \circ e$.
The maps $s$ and $t$ have components. The equality has to hold for all
components and we get
\begin{align*}
  s \circ e &= t \circ e\\
  s_n \circ e &= t_n \circ e, \quad \forall\ K \overset{n}{\rightarrow} J \in \bm I \\
  Dn \circ p_K \circ e &= p_J \circ e
\end{align*}
With this setup we show that we have a universal cone over $D$.
We take $E$ as the vertex and
$\uprho_I \coloneqq p_I \circ e, \forall\ I \in \bm I$
as the projection maps of the cone.
Clearly, we have all necessary projection maps and we only need to verify
the commuting condition of cones.
We have all maps $K \overset{n}{\rightarrow} J \in \bm I$
\begin{align*}
  Dn \circ p_K \circ e &= p_J \circ e \\
  Dn \circ \uprho_K &= \uprho_J
\end{align*}
which corresponds to the commuting diagram
\[
  \begin{tikzcd}
    E \arrow{r}{\uprho_K} \arrow{dr}[swap]{\uprho_J} & D_K \arrow{d}{Dn}\\
    & D_J
  \end{tikzcd}
\]
This is exactly the commuting condition we need for a cone.\\
Since any other cone over $D$
is also a fork of $s$ and $t$
it follows that the equalizer $E$ is also a universal cone,
that is, a limit of $D$.
\end{proof}
\begin{remark}
  If $\mathscr C$ has all products including infinite products then
  this construction works even for infinite diagrams.
  If $\mathscr C$ only has finite products, then we can
  use this construction for
  limits of finite diagrams.
\end{remark}

To summarize, we can construct arbitrary limits in the following way.
If $\mathscr C$ has all products and equalizers then we can
construct all limits by
\hyperref[arbitrarylimits]{Theorem \ref*{arbitrarylimits}}.
Equalizers can be constructed using finite products and pullbacks
by \hyperref[arbitrarylimits]{Theorem \ref*{pullbackequalizer}}.
In the case that we want to construct infinite limits
the existence of infinite products is necessary
for which we have no construction.
If we want to construct finite limits, then finite products are sufficient.
Finite products can be constructed from binary products and a terminal object
by \hyperref[finiteproducts]{Lemma \ref*{finiteproducts}}.
Binary products can in turn be constructed from a pullback and a terminal object
by \hyperref[arbitrarylimits]{Theorem \ref*{pullbackproduct}}.
Having pullbacks and a terminal object is therefore the most basic requirement from
which it is possible to bootstrap the construction of arbitrary finite limits.

\section{Colimits}
 Colimits are the dual of limits.
The motivation to dualize limits is to obtain dualized versions of theorems for limits.
Many concepts in mathematics turn out to be colimits and it is
therefore useful to have a general definition for them as well.
\begin{definition}
  A \textbf{colimit} is a universal cocone over a diagram.
\end{definition}
A cocone is the dualized version of a cone.
Recall that a cone has maps from the vertex into the diagram. Dually,
a cocone has maps from the diagram into the vertex.
\begin{definition}
  Let $D : \bm I \to \mathscr C$ be a diagram.
  A \textbf{cocone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $D_I \overset{ f_I}{\rightarrow} U,\ \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      D_I \arrow[rr, "D_m"]\arrow[dr, swap, "f_I"] & & D_J \arrow{dl}{f_J}\\
      & U  &
    \end{tikzcd}
  \]
  That is $f_I = f_J \circ D_m$.
\end{definition}
The universality condition is also dualized as expected.
For limits the condition is the existence of a unique map from
any other cone to the limit cone.
For colimits the condition is therefore dually the existence of a unique map
from the colimit to any other cocone.
\begin{definition}
  A cocone $(D_I\overset{f_I}{\to} U)_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cocones $(D_I \overset{g_I}{\to} V)_{I\in \bm{I}}$ there is a unique
  map $h \from U \to V$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      D_I \arrow[dr, swap, "g_I"] \arrow{r}{f_I} & U \arrow{d}{h} \\
      & V
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}
\begin{example}
\[
    \begin{tikzcd}
      & D_K \arrow{d}{D_n}
      \arrow[bend left=30,color=blue]{ddr}{f_K}
      \arrow[bend left=50,color=red]{dddrr}{g_K} & & \\
      D_I \arrow{r}{D_m}
      \arrow[bend right=30,color=blue]{drr}{f_I}
      \arrow[bend right=40,color=red]{ddrrr}{g_I} &
      D_J \arrow[bend left=20,swap, color=blue]{dr}{f_J}
      \arrow[bend left=40,color=red]{ddrr}{g_J} & & \\
      & & {\color{blue}U} \arrow[dashed]{dr}{h}& \\
      & & & {\color{red}V}
    \end{tikzcd}
  \]
\end{example}
